---
title: "03-Data-Preparation-Network-Data"
---

In this notebook I pre-process the network data.

I downloaded the following data from [Hamilton Open Data Portal](https://open.hamilton.ca/datasets/370a19ad2e1543989e6dcb00254b3eb4_3):

- A shape file with the neighborhood boundaries: Neighbourhood_Boundaries.shp

Further, the notebook downloads Open Street Maps information about the network.

## Preliminaries

Load the packages used in this notebook:
```{r message=FALSE}
library(sf)
library(tidyverse)
library(units)
```

## Neighborhoods

Read neighborhoods file:
```{r}
hamilton_neighborhoods <- st_read("Neighbourhood_Boundaries.shp")
```

Check that the projection matches the taz boundaries:
```{r}
st_crs(hamilton_neighborhoods)
```

Change projection of object:
```{r}
hamilton_neighborhoods <- hamilton_neighborhoods %>%
  st_transform(crs = 32617)
```

Plot neighborhoods:
```{r}
ggplot(data = hamilton_neighborhoods) + geom_sf() + geom_sf_text(aes(label = NH_NAME), size = 2)
```

Rename columns:
```{r}
hamilton_neighborhoods <- hamilton_neighborhoods %>%
  rename(Name = NH_NAME, Number = NH_NUMBER)
```

## Save processed data

Save neighborhoods:
```{r}
usethis::use_data(hamilton_neighborhoods, 
                  overwrite = TRUE)
```
