---
title: "05-Data-Preparation-OSM-Opportunities"
---

In this notebook I pre-process data for urban opportunities (e.g., restaurants/eateries).

The notebook downloads Open Street Maps information about the network.

## Preliminaries

Load the packages used in this notebook:
```{r message=FALSE}
library(osmdata)
library(sf)
library(tidyverse)
library(units)
```

## Traffic Analysis Zones

Read the boundaries of the traffic analysis zones:
```{r}
hamilton_taz <- st_read("Hamilton-CMA-tts06.shp")
```

Project object:
```{r}
hamilton_taz <- hamilton_taz %>%
  st_transform(crs = 32617)
```

Prepare traffic analysis zones:
```{r}
hamilton_taz <- hamilton_taz %>%
  transmute(ID,
            AREA = st_area(geometry) %>% set_units(km^2) %>% drop_units(),
            GTA06 = as.character(GTA06))
```

## OSM - eateries

Obtain a bounding box for Hamilton:
```{r}
#gtha_bb <- st_bbox(taz_boundaries)
hamilton_bb <- hamilton_taz %>%
  st_transform(4326) %>%
  st_bbox()
```

Use the bounding box to retrieve OSM _restaurant_ data for Hamilton:
```{r}
hamilton_bar_osm <- osmdata::opq(bbox =  hamilton_bb) %>% 
  osmdata::add_osm_feature(key = "amenity",
                           value = "bar") %>% 
  osmdata::osmdata_sf()

hamilton_bars <- hamilton_bar_osm$osm_points %>% 
  transmute(osm_id, name, amenity, type = "Bar") %>%
  st_transform(crs= 32617)

hamilton_cafe_osm <- osmdata::opq(bbox =  hamilton_bb) %>% 
  osmdata::add_osm_feature(key = "amenity",
                           value = "cafe") %>% 
  osmdata::osmdata_sf()

hamilton_cafes <- hamilton_cafe_osm$osm_points %>% 
  transmute(osm_id, name, amenity, type = "Cafe") %>%
  st_transform(crs= 32617)

hamilton_fast_food_osm <- osmdata::opq(bbox =  hamilton_bb) %>% 
  osmdata::add_osm_feature(key = "amenity",
                           value = "fast_food") %>% 
  osmdata::osmdata_sf()

hamilton_fast_food <- hamilton_fast_food_osm$osm_points %>% 
  transmute(osm_id, name, amenity, type = "Fast Food") %>%
  st_transform(crs= 32617)

hamilton_food_court_osm <- osmdata::opq(bbox =  hamilton_bb) %>% 
  osmdata::add_osm_feature(key = "amenity",
                           value = "food_court") %>% 
  osmdata::osmdata_sf()

hamilton_food_court <- hamilton_food_court_osm$osm_points %>% 
  transmute(osm_id, name, amenity, type = "Food Court") %>%
  st_transform(crs= 32617)

hamilton_pub_osm <- osmdata::opq(bbox =  hamilton_bb) %>% 
  osmdata::add_osm_feature(key = "amenity",
                           value = "pub") %>% 
  osmdata::osmdata_sf()

hamilton_pubs <- hamilton_pub_osm$osm_points %>% 
  transmute(osm_id, name, amenity, type = "Pub") %>%
  st_transform(crs= 32617)

hamilton_restaurant_osm <- osmdata::opq(bbox =  hamilton_bb) %>% 
  osmdata::add_osm_feature(key = "amenity",
                           value = "restaurant") %>% 
  osmdata::osmdata_sf()

hamilton_restaurants <- hamilton_restaurant_osm$osm_points %>% 
  transmute(osm_id, name, amenity, type = "Restaurant") %>%
  st_transform(crs= 32617)
```

Join all types of eateries:
```{r}
hamilton_sustenance <- rbind(hamilton_bars, 
                           hamilton_cafes,
                           hamilton_fast_food,
                           hamilton_food_court,
                           hamilton_pubs,
                           hamilton_restaurants)
```

Plot eateries:
```{r}
ggplot(hamilton_sustenance) +
  geom_sf(aes(color = type))
```

## Save processed data

Save Hamilton eateries:
```{r}
usethis::use_data(hamilton_sustenance,
                  overwrite = TRUE)
```

